Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/introduction.tex
update fac-overview and unhack code font size
tweak fac-overview
done interactive
misc edits
Wes comments. minor wording/typo issues checked in directly, others left as "WRW" comments (search) that I will mention via email.

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-23:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\begin{code}
let rec fac n =
  if n <= 0 then
    true
  else
    n * fac (n-1)
\end{code}
\end{minipage}
\hspace{0.3in}
\begin{minipage}{.3\linewidth}
  \includegraphics[height=2.5in]{fac-overview.png}
\end{minipage}
\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}}
\label{fig:factorial}
\end{figure}

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 3, characters 8-23:
 Error: This expression has type bool
        but an expression was expected
        of type int
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{lerner_searching_2007,chen_counter-factual_2014,zhang_toward_2014,pavlinovic_finding_2014},
but an accurate source location still does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\centering
\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\end{minipage}
% \hspace{0.3in}
\begin{minipage}{.49\linewidth}
\centering
  \includegraphics[height=1.5in]{fac-overview.png}
\end{minipage}
\caption{(L) An ill-typed \texttt{fac} function; (R) Dynamically witnessing the type error in \texttt{fac}.}
\label{fig:factorial}
\end{figure}
